Arcus X-ray telescope performance predictions and alignment requirements

### 2024-05-20 papers/special_JATIS_tolerances/tolerances.tex
submitted paper to JATIS

@@

\author[a]{Hans Moritz G\"unther$^*$}
\author[b]{Peter Cheimets}
\author[c]{Casey T. DeRoo}
\author[a,d]{Ralf K. Heilmann}
\author[a]{Eric D. Miller}
\author[e]{Andrew Ptak}
\author[b]{Randall K. Smith}

\affil[a]{MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
\affil[b]{Center for Astrophysics, Harvard-Smithsonian Astrophysical Observatory, Cambridge, MA 02138, USA}
\affil[c]{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
\affil[d]{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
\affil[e]{NASA Goddard Space Flight Center, Greenbelt, MD 20771, USA}

@@

\begin{abstract}
Arcus is a concept for a probe class mission to deliver high-resolution FUV and X-ray spectroscopy with two separate instruments. The X-ray spectrograph (XRS) is the focus of this paper. It consists of four spectral channels arranged in a double-tilted Rowland torus geometry.  It combines cost-effective silicon pore optics (SPO) with high-throughput critical-angle transmission (CAT) gratings to achieve $R> 3000$ in a bandpass from 12-50 \AA. We present ray-tracing studies to derive performance characteristics such as the spectral resolving power and effective area, and look at the best positioning of the four channels to improve the resiliency towards misalignments and reduce the overall impact of chip gaps. We study the effect of misalignments on the performance, and present alignment requirements in six degrees of freedom for all optical elements in the XRS. We
conclude that most tolerances can be achieved with mechanical means alone.
\end{abstract}

@@

% FOR SUBMISSION
% \begin{spacing}{2}   % use double spacing for rest of manuscript

% For my own while writing. It's just easier to read this way
\begin{spacing}{1}

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
High-resolution X-ray and UV spectroscopy open a window into the physics of the universe that often cannot be observed with any other technique. X-rays and UV can probe the hottest and most ionized gas that remains invisible in longer wavelengths because the high ionization levels do not produce observable transitions in the radio, infrared or optical. Thus, observations in the UV and X-rays probe the most energetic processes in a number of systems. A few examples are: stellar space weather \cite{2022AN....34320019B}, the accretion onto young stars, where the UV and X-rays come from the energetic infall of material from the disk onto the stellar surface; the innermost regions of the accretion disks around black holes; and the absorption of background (X-ray) light from distant, bright continuum sources by the warm-hot inter-galactic medium to find the missing baryons\cite{10.1117/12.2231193}. High-resolution UV and X-ray spectroscopy that resolves the profiles of individual emission or absorption lines is particularly valuable, because it allows us to address a host of physical questions that cannot be answered by simply measuring the broad-band X-ray flux.
To follow-up with one specific example: in the case of accretion onto young stars, resolving the kinematic line profile of emission in different ions can tell us which part of the emission is formed in the infalling, red-shifted accretion column, which part is related to the stellar corona seen at the rest wavelength of the star (possibly with small velocities from up and downflows in the corona), and which part is formed in (or possibly absorbed by) the blue-shifted outflow.

@@

Beginning in the very early phases of the mission design we performed
ray-tracing to verify the performance characteristics of Arcus and to
improve the design\cite{10.1117/12.2232157}; in particular in \citenum{10.1117/12.2312678} we utilized ray-tracing to determine the allowable misalignments of the mechanical elements that make up Arcus. Ray-tracing is particularly well suited for the problem, because it allows an arbitrary misalignment to be introduced for any element in the optical path and it can quickly iterate over different input parameters. Since the work in \citenum{10.1117/12.2312678}, the Arcus concept has evolved yet again. In its most recent version, Arcus is a NASA Probe-class mission with significantly increased effective area and the addition of the UV instrument. Since the UV telescope has a smaller field-of-view than the X-ray instrument, it will define the fiducial pointing direction; now misalignment between the two separate instruments is a new concern.

@@

\subsection{Silicon Pore Optics (SPOs)}
The SPOs are a technology developed and matured for Athena. They provide a large effective area at a relatively low weight and cost\cite{10.1117/12.2188988,10.1117/12.2599339,10.1117/12.2677388}. In Arcus, the SPO petals use ``sub-aperturing'', which means that the SPOs do not cover a full circle, but only a narrow wedge. That provides Arcus with an asymmetric PSF that is only about 1.5\ arcsec wide in the dispersion direction. It also means that the individual petals where the SPOs are mounted are not circular, but roughly rectangular, and four separate panels (one per optical channel) can fit into the front assembly of the telescope; see Figure~\ref{fig:arcus} for a rendering.

\subsection{Critial-Angle Transmission (CAT) gratings}
Critical-angle transmission (CAT) gratings are manufactured at the Space Nanotechnology Labo\-ra\-tory. Progress on the development and manufacturing of the CAT gratings is given in a series of papers stretching back many years\cite{10.1117/12.926827,2022ApJ...934..171H}.
CAT gratings have a very high aspect ratio, where the grating bars are about 50 times higher than they are wide. Arcus will feature CAT grating with a period of 200~nm in the dispersion direction, 140~nm gaps between the bars and a depth of 5.6~$\mu$m. The gratings are manufactured with sizes up to $32\times32$ mm$^2$. Individual grating bars are held in place by a perpendicular support structure of wider Si bars (L1 support) and a hexagonal frame (L2 support) that the gratings bars and L1 supports rest on. The L2 mesh blocks $\sim 19$\% of the area, and the recently fabricated L1 meshes block between 10 and 18\% -- see Fig.~\ref{fig:cell} for an image and a schematic.

\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[height=5cm]{unitcell+pic.jpg}
    \end{tabular}
    \end{center}
    \caption {\label{fig:cell}
    Left: Schematic depiction of the structural hierarchy of the CAT grating membrane (see text). Right: Photograph of a back-illuminated $32\times32$ mm$^2$ prototype CAT grating membrane, showing the hexagonal L2 mesh and optical diffraction from the L1 mesh.
    }
\end{figure}

@@

\subsection{Double-tilted Rowland Torus layout}
Each channel follows a Rowland torus layout\cite{Beuermann:78} where the gratings are mounted on the surface of the torus and the detectors on the opposite side. Detectors follow the curved Rowland circle, which optimizes the width in the dispersion direction to increase the spectral resolving power of the instrument at the cost of a slightly wider photon distribution in cross-dispersion direction. For each channel, the Rowland torus is tilted by 3.6175~deg, about twice the blaze angle\cite{10.1117/12.856482,10.1117/12.2273011,DTRS}. This number is set such that the distance between two channels matches the space required to mount the grating petals next to each other. At the same time, all channels can be imaged on the same set of detectors (Fig.~\ref{fig:geometry_camera}).
The different tilted tori are arranged in such a way, that they overlap in the Rowland circle, so that the optimal detector position for one channel is also the optimal detector position for all other channels.
Two cameras are positioned to capture the zeroth order on one side and the blaze peak on the other with a gap in between where so little signal is found that it is not useful to capture it with CCDs.

\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[width=\textwidth]{geometry_camera}
    \end{tabular}
    \end{center}
    \caption {\label{fig:geometry_camera}
    Layout of the CCDs (orange rectangles) in the focal plane. Given the resolution of the figure, most chip gaps are not visible. Also note, that the x and the y axes are scaled differently (Figure~\ref{fig:4} shows correct aspect of the CCDS, but not the correct gap between the two detector arrays). Two cameras with eight CCDs each are located in the focal plane. In this plot, 0 is the geometric center of all the Rowland tori involved. Note that the cameras are intentionally not symmetric to 0.
    The position of the four optical axes for the four channels is marked with a blue pentagon.  The dispersion ($x$ - green horizontal arrows) and the cross-dispersion ($y$ - blue vertical arrows) of the different spectra are parallel to each other and point in the same direction. Two channels disperse left-to-right, and two right-to-left.
    }
\end{figure}

@@

Each of the four channels of the XRS has a separate zero-order position and disperses the light to a different location on the detector. This has two main advantages: there is no need to align all channels to the same zero-order position within the width of the point-spread function (PSF, less than 1.5 arcsec in dispersion direction), and any wavelength of light that falls in a chip gap for one channel will be observable in the others; see Figure~\ref{fig:geometry_camera}. At the same time, this layout allows us to use the same detectors for all four channels, which reduces cost and complexity. More details of the optical layout are given in our previous ray-tracing papers\cite{10.1117/12.2273011,10.1117/12.2312678,DTRS}.

\section{Ray-traces}
We study the performance and alignment requirements of Arcus with ray-tracing. We use the ray-tracing code MARXS\cite{marxs1.1,2017AJ....154..243G}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with a development version with git hash f4b68dce0957 to include Arcus specific code. MARXS has been used for several mission proposals and is backed by hundreds of units tests and verified against laboratory works and on-orbit Chandra observations where feasible. MARXS does not calculate material properties such as reflectivities or grating efficiencies; instead they are read in from tables that can be based on laboratory data or simulations calculated by other programs.

@@

After running the baseline version, one element of Arcus is shifted in one degree of freedom, e.g.\ all CCDs are shifted by 1~mm in the dispersion direction. The ray-trace is repeated, again $R$ and $A_\mathrm{eff}$ are calculated, then all CCDs are shifted by 2~mm and so on. After testing out the parameter space in dispersion direction, the CCDs will be shifted in cross-dispersion direction. In this way, each element (for example the CCD array), will be misaligned by various amounts in one of 6 degrees of freedom (shift along $x$, $y$, $z$, and rotation around $x$, $y$, $z$).
In the Arcus coordinate system, the $z$-axis is parallel to the optical axis, the $x$-axis is the grating dispersion direction and the $y$-axis is the cross-dispersion direction. Unless noted, rotations are not done around the origin of the coordinate system, but around the center of an element (e.g.\ the center of a SPO petal).

In the first stage, only one degree of freedom is changed at a time, because it is not computationally feasible to explore the entire parameter space. From those results, we identify where the alignment is easily (e.g.\ just from simple machining tolerances) much better than the requirements. In a second step, we can then run ray-traces where all degrees of freedom are varied according to the misalignment budget and thus check if the assumptions going into combining the misalignments in different degrees of freedom hold or if non-linear interactions degrade $R$ and $A_\mathrm{eff}$ more than expected. In this case, we would have to revise the misalignment budget appropriately.

@@

Figure~\ref{fig:CAT_global} shows the effect of a change in the CAT grating petal position, while the SPO petal and the cameras are fixed. Rotations around either $x$ or $y$ mean that the CAT gratings on one side move up, while the other side moves down, changing the path length of the diffracted photons. Those photons coming from the high CAT gratings travel further along the dispersion direction than those from the low gratings, thus causing the dispersed spot to smear out, which reduces $R$. These rotation need to be kept at the level of a few arcminutes. Rotation around $z$ changes the direction of the dispersed light and, for large angles, the dispersed orders miss the CCD.

Figure~\ref{fig:CAT_individual} examines the rotations of individual gratings. For each grating in the CAT petal, a different misalignment is drawn from a normal distribution. Rotations around $x$ (the dispersion direction, see panels a and d in the figure) have little effect on $R$. Since the facets have finite size, their edges differ from the Rowland torus by a little already, and adding a little more rotation does not change much. Rotations in $y$ quickly reduce $A_\mathrm{eff}$ though, because the CCDs are placed for a certain blaze peak. Rotating the CAT gratings shifts the blaze peak and photons miss the CCDs. Rotation around $z$ causes more and more signal to miss the detector. There is no sharp cut-off, because the CAT gratings in the simulation have a distribution of rotation angles, and the larger the Gaussian $\sigma$ is, the more CAT gratings will send their dispersed photons to positions where they cannot be detected. From this figure, we can determine that the alignment tolerance for rotation around $x$ can be large, while the other two directions are of order a few arcminutes to prevent significant degradation of the Arcus performance.

Figure~\ref{fig:detector_global} shows simulations for translating the detector with respect to the forward assembly (SPOs and CAT gratings). The most important degree of freedom is a change in focus (panel c). A shift along $y$ has no impact, as long as it is small enough to keep the dispersed spectrum on the CCDs. For the particular channel simulated here the spectrum drops off the CCDs for $y$ shifts for about -15~mm on one side and about +5~mm on the other side. A shift of 5~mm or more will drop at least one channel of the detector.

@@

We run simulations for rotations and translations for each possible mechanical misalignment in Arcus, as well as for a few other parameters such as the SPO mirror PSF, the repeatability of the CAT grating period, the surface flatness of the CAT gratings etc, but it is beyond the scope of this article to show each result in a figure. Inspecting all results, and considering how easy or hard it is to improve alignments in each degree of freedom, we set the alignment tolerances listed in Table~\ref{tab:tolerances}.

\begin{table}
    \caption{Arcus alignment tolerances ($1\sigma$)\label{tab:tolerances}}
    \centering
    \begin{tabular}{l|cccccc}
        \hline\hline
    alignment & trans x & trans y & trans z & rot x & rot y & rot z \\
     & $\mathrm{mm}$ & $\mathrm{mm}$ & $\mathrm{mm}$ & $\mathrm{{}^{\prime\prime}}$ & $\mathrm{{}^{\prime\prime}}$ & $\mathrm{{}^{\prime\prime}}$ \\
     \hline
    individual SPO in petal & 0.007 & 0.033 & 0.167 & 100 & 100 & 3 \\
    CAT petal to SPO petal & 0.333 & 0.333 & 0.333 & 100 & 100 & 200 \\
    CAT windows to CAT petal & 0.333 & 0.333 & 0.067 & 100 & 60 & 100 \\
    individual CAT to window & 0.333 & 0.333 & 0.067 & 100 & 60 & 100 \\
    Camera to front assembly & 1.667 & 0.667 & 0.333 & 60 & 60 & 60 \\
    \end{tabular}
\end{table}

@@

\section{Future Work}
At this point, the main limitation of this work is our simplified mirror model. We expect it to work well for on-axis sources, but it does not reproduce the vignetting which becomes significant at about 2~arcmin. Our simplified mirror model also leads to a flat focal plane, while a Wolter-Type II optic typically has a slightly curved focal surface. Detailed ray-trace models specific to the Athena SPOs are available\cite{10.1117/12.2232230,10.1117/12.2594461,10.1117/12.2628133} and can be coupled with our code for CAT gratings and detectors.

@@

\subsection* {Code, Data, and Materials Availability}
We use the ray-tracing code MARXS\cite{marxs1.1,2017AJ....154..243G}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with a development version with git hash f4b68dce0957. The database of efficiencies for gratings is considered a trade secret by the Space Nanotechnology Laboratory and cannot be shared in open repositories - contact co-author Ralf Heilmann for access.

@@

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Peter Cheimets}
Please send me your biography (approx 75 words).

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Casey DeRoo}
Please send me your biography (approx 75 words).

@@

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Eric D. Miller}
Please send me your biography (approx 75 words).

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Andrew Ptak}
Please send me your biography (approx 75 words).

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Randall K. Smith}
Please send me your biography (approx 75 words).

>>>>

\usepackage{lineno}
\linenumbers

@@

\author[a]{Hans Moritz G\"unther$^*$}
\author[b]{Peter Cheimets}
\author[c]{Casey T. DeRoo}
\author[a,d]{Ralf K. Heilmann}

\affil[a]{MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
\affil[b]{Center for Astrophysics, Harvard-Smithsonian Astrophysical Observatory, Cambridge, MA 02138, USA}
\affil[c]{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
\affil[d]{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}

@@

\begin{abstract}
Arcus is a concept for a Probe-class X-ray mission to deliver high-resolution FUV and X-ray spectroscopy with two separate instruments. The X-ray spectrograph (XRS) is the focus of this paper. It consists of four spectral channels arranged in a double-tilted Rowland torus geometry.  It combines cost-effective silicon pore optics (SPO) with high-throughput critical-angle transmission (CAT) gratings to achieve $R> 3000$ in a bandpass from 12-50 \AA. We present ray-tracing studies to derive performance characteristics such as the spectral resolving power and effective area, and look at the best positioning of the four channels to improve the resiliency towards misalignments and reduce the overall impact of chip gaps. We study the effect of misalignments on the performance, and present alignment requirements in six degrees of freedom for all optical elements in the XRS. We
conclude that most tolerances can be achieved with mechanical means alone.
\end{abstract}

@@

% FOR SUBMISSION
\begin{spacing}{2}   % use double spacing for rest of manuscript

% For my own while writing. It's just easier to read this way
% \begin{spacing}{1}

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
High-resolution X-ray and UV spectroscopy open a window into the physics of the universe that often cannot be observed with any other technique. X-rays and UV can probe the hottest and most ionized gas that remains invisible in longer wavelengths because the high ionization levels do not produce observable transitions in the radio, infrared or optical. Thus, observations in the UV and X-rays probe the most energetic processes in a number of systems. A few examples are: stellar space weather \cite{2022AN....34320019B}, the accretion onto young stars, where the UV and X-rays come from the energetic infall of material from the disk onto the stellar surface; the innermost regions of the accretion disks around black holes; and the absorption of background (X-ray) light from distant, bright continuum sources by the warm-hot inter-galactic medium to find the missing baryons\cite{10.1117/12.2231193,10.1117/12.2529499}. High-resolution UV and X-ray spectroscopy that resolves the profiles of individual emission or absorption lines is particularly valuable, because it allows us to address a host of physical questions that cannot be answered by simply measuring the broad-band X-ray flux.
To follow-up with one specific example: in the case of accretion onto young stars, resolving the kinematic line profile of emission in different ions can tell us which part of the emission is formed in the infalling, red-shifted accretion column, which part is related to the stellar corona seen at the rest wavelength of the star (possibly with small velocities from up and downflows in the corona), and which part is formed in (or possibly absorbed by) the blue-shifted outflow \cite{10.1117/12.2529499}.

@@

Beginning in the very early phases of the mission design we performed
ray-tracing to verify the performance characteristics of Arcus and to
improve the design\cite{10.1117/12.2232157}; in particular in \citenum{10.1117/12.2312678,10.1117/12.2677455} we utilized ray-tracing to determine the allowable misalignments of the mechanical elements that make up Arcus. Ray-tracing is particularly well suited for the problem, because it allows an arbitrary misalignment to be introduced for any element in the optical path and it can quickly iterate over different input parameters. Since the work in \citenum{10.1117/12.2312678}, the Arcus concept has evolved yet again.

@@

\subsection{Silicon Pore Optics (SPOs)}
The SPOs are a technology developed and matured for Athena. They provide a large effective area at a relatively low weight and cost\cite{10.1117/12.2188988,10.1117/12.2599339,10.1117/12.2677388}. In Arcus, the SPO petals use ``sub-aperturing'', which means that the SPO do not cover a full circle, but only a narrow wedge. That provides Arcus with an asymmetric point-spread function (PSF) that is only about 1.5\ arcsec wide in the dispersion direction. It also means that the individual petals where the SPOs are mounted are not circular, but roughly rectangular, and four separate panels (one per optical channel) can fit into the front assembly of the telescope; see Figure~\ref{fig:arcus} for a rendering.

\subsection{Critial-Angle Transmission (CAT) gratings}
Critical-angle transmission (CAT) gratings are manufactured at the Space Nanotechnology Labo\-ra\-tory.
Progress on the development, manufacturing and x-ray performance of CAT gratings is given in a series of papers stretching back many years\cite{OE2008,AO2011,EIPBN2016,AO2019,2022ApJ...934..171H}.
CAT gratings have a very high aspect ratio, where the grating bars are about 50 times higher than they are wide. Arcus will feature CAT grating with a period of 200~nm in the dispersion direction, 140~nm gaps between the bars and a depth of 5.6~$\mu$m. The gratings are manufactured with sizes up to $32\times32$ mm$^2$. Individual grating bars are held in place by a perpendicular support structure of wider Si bars (L1 support) and a hexagonal frame (L2 support) that the gratings bars and L1 supports rest on. The L2 mesh blocks $\sim 19$\% of the area, and the recently fabricated L1 meshes block between 10 and 18\% -- see Fig.~\ref{fig:cell} for an image and a schematic.

\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[height=5cm]{unitcell+pic.png}
    \end{tabular}
    \end{center}
    \caption {\label{fig:cell}
    Left: Schematic depiction of the structural hierarchy of the CAT grating membrane (see text). Right: Photograph of a back-illuminated $32\times32$ mm$^2$ prototype CAT grating membrane, showing the hexagonal L2 mesh and optical diffraction from the L1 mesh.
    }
\end{figure}

@@

\subsection{Double-tilted Rowland Torus layout}
Each channel follows a Rowland torus layout\cite{Beuermann:78} where the gratings are mounted on the surface of the torus and the detectors on the opposite side. Detectors follow the curved Rowland circle, which optimizes the width in the dispersion direction to increase the spectral resolving power of the instrument at the cost of a slightly wider photon distribution in cross-dispersion direction. For each channel, the Rowland torus is tilted by 3.6175~deg, about twice the blaze angle\cite{10.1117/12.856482,10.1117/12.2273011,DTRS}. This number is set such that the distance between two channels matches the space required to mount the grating petals next to each other. At the same time, all channels can be imaged on the same set of detectors (Fig.~\ref{fig:geometry_camera}).
The different tilted tori are arranged in such a way, that they overlap in the Rowland circle, so that the optimal detector position for one channel is also the optimal detector position for all other channels.
Two cameras are positioned to capture the zeroth order on one side and the blaze peak on the other. There is a gap between the two cameras because so little signal is found in that region that it is not useful to capture it with CCDs.

\begin{figure} [ht]
    \begin{center}
    \begin{tabular}{c} %% tabular useful for creating an array of images
    \includegraphics[width=\textwidth]{geometry_camera}
    \end{tabular}
    \end{center}
    \caption {\label{fig:geometry_camera}
    Layout of the CCDs (orange rectangles) in the focal plane. Given the resolution of the figure, most chip gaps are not visible. Also note, that the x and the y axes are scaled differently. Two cameras with eight CCDs each are located in the focal plane. In this plot, 0 is the geometric center of all the Rowland tori involved. Note that the cameras are intentionally not symmetric to 0.
    The position of the four optical axes for the four channels is marked with a blue pentagon.  The dispersion ($x$ - green horizontal arrows) and the cross-dispersion ($y$ - blue vertical arrows) of the different spectra are parallel to each other and point in the same direction. Two channels disperse left-to-right, and two right-to-left.
    }
\end{figure}

@@

Each of the four channels of the XRS has a separate zero-order position and disperses the light to a different location on the detector. This has two main advantages: there is no need to align all channels to the same zero-order position within the width of the PSF (less than 1.5 arcsec in dispersion direction), and any wavelength of light that falls in a chip gap for one channel will be observable in the others; see Figure~\ref{fig:geometry_camera}. At the same time, this layout allows us to use the same detectors for all four channels, which reduces cost and complexity. More details of the optical layout are given in our previous ray-tracing papers\cite{10.1117/12.2273011,10.1117/12.2312678,DTRS}.

\section{Ray-traces}
We study the performance and alignment requirements of Arcus with ray-tracing. We use the ray-tracing code MARXS\cite{2017AJ....154..243G,marxs2.0}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with version 2.0. MARXS has been used for several mission proposals and is backed by hundreds of units tests and verified against laboratory works and on-orbit Chandra observations where feasible. MARXS does not calculate material properties such as reflectivities or grating efficiencies; instead they are read in from tables that can be based on laboratory data or simulations calculated by other programs.

@@

After running the baseline version, one element of Arcus is shifted in one degree of freedom, e.g.\ all CCDs are shifted by 1~mm in the dispersion direction. The ray-trace is repeated, again $R$ and $A_\mathrm{eff}$ are calculated, then all CCDs are shifted by 2~mm and so on. After testing out the parameter space in dispersion direction, the CCDs will be shifted in cross-dispersion direction. In this way, each element (for example, the CCD array), will be misaligned by various amounts in one of 6 degrees of freedom (shift along $x$, $y$, $z$, and rotation around $x$, $y$, $z$).
In the Arcus coordinate system, the $z$-axis is parallel to the optical axis, the $x$-axis is the grating dispersion direction and the $y$-axis is the cross-dispersion direction. Unless noted, rotations are not done around the origin of the coordinate system, but around the center of an element (e.g.\ the center of a SPO petal).

In the first stage, only one degree of freedom is changed at a time, because it is not computationally feasible to explore the entire parameter space. From those results, we identify where the alignment is easily (e.g.\ just from simple machining tolerances) much better than the requirements. In a second step, we can then run ray-traces where all degrees of freedom are varied according to the misalignment budget and thus check if the assumptions going into combining the misalignments in different degrees of freedom hold or if non-linear interactions degrade $R$ and $A_\mathrm{eff}$ more than expected. Should this be the case, we revise the misalignment budget appropriately.

@@

Figure~\ref{fig:CAT_global} shows the effect of a change in the CAT grating petal position, while the SPO petal and the cameras are fixed. Rotations around either $x$ or $y$ mean that the CAT gratings on one side move up, while the other side moves down, changing the path length of the diffracted photons. Those photons coming from the high CAT gratings travel further along the dispersion direction than those from the low gratings, thus causing the dispersed spot to smear out, which reduces $R$. These rotations need to be kept at the level of a few arcminutes. Rotation around $z$ changes the direction of the dispersed light and, for large angles, the dispersed orders miss the CCD.

Figure~\ref{fig:CAT_individual} examines the rotations of individual gratings. For each grating in the CAT petal, a different misalignment is drawn from a normal distribution. Rotations around $x$ (the dispersion direction, see panels a and d in the figure) have little effect on $R$. Since the flat facets have finite size, their edges differ from the Rowland torus by a little already, and adding a little more rotation does not change much. Rotations in $y$ quickly reduce $A_\mathrm{eff}$ though, because the CCDs are placed for a certain blaze peak. Rotating the CAT gratings shifts the blaze peak and photons miss the CCDs. Rotation around $z$ causes more and more signal to miss the detector. There is no sharp cut-off, because the CAT gratings in the simulation have a distribution of rotation angles, and the larger the Gaussian $\sigma$ is, the more CAT gratings will send their dispersed photons to positions where they cannot be detected. From this figure, we can determine that the alignment tolerance for rotation around $x$ can be large, while the other two directions are of order a few arcminutes to prevent significant degradation of the Arcus performance.

Figure~\ref{fig:detector_global} shows simulations for translating the detector with respect to the forward assembly (SPOs and CAT gratings). The most important degree of freedom is a change in focus (panel c). A shift along $y$ has no impact, as long as it is small enough to keep the dispersed spectrum on the CCDs. For the particular channel simulated here, the spectrum drops off the CCDs for $y$ shifts for about -15~mm on one side and about +5~mm on the other side. A shift of 5~mm or more will drop at least one channel of the detector.

@@

We run simulations for rotations and translations for each possible mechanical misalignment in Arcus, as well as for a few other parameters such as the SPO mirror PSF, the repeatability of the CAT grating period, the surface flatness of the CAT gratings etc. Inspecting all results, and considering how easy or hard it is to improve alignments in each degree of freedom, we set the alignment tolerances listed in Table~\ref{tab:tolerances}.

\begin{table}
    \caption{Arcus alignment tolerances ($1\sigma$)\label{tab:tolerances}}
    \centering
    \begin{tabular}{l|cccccc}
        \hline\hline
    alignment & trans x & trans y & trans z & rot x & rot y & rot z \\
     & $\mathrm{mm}$ & $\mathrm{mm}$ & $\mathrm{mm}$ & $\mathrm{{}^{\prime\prime}}$ & $\mathrm{{}^{\prime\prime}}$ & $\mathrm{{}^{\prime\prime}}$ \\
     \hline
    individual SPO in petal & 0.007 & 0.033 & 0.167 & 100 & 100 & 3 \\
    CAT petal to SPO petal & 0.333 & 0.333 & 0.333 & 100 & 100 & 200 \\
    CAT windows to CAT petal & 0.333 & 0.333 & 0.067 & 100 & 60 & 100 \\
    individual CAT to window & 0.333 & 0.333 & 0.067 & 100 & 60 & 100 \\
    Camera to front assembly & 1.667 & 0.667 & 0.333 & 60 & 60 & 200 \\
    \end{tabular}
\end{table}

@@

We use the feedback we gain from our ray traces to revisit the error budget and look for places where we can relax it without impacting performance.
For example, we have been able to relaxe rotational alignment tolerances for CAT grating membranes and/or facets to their grating window by a factor of three relative to Table 1 with acceptable minimal impact on XRS performance.

\section{Future Work}
At this point, the main limitation of this work is our simplified mirror model. We expect it to work well for on-axis sources, but it does not reproduce the vignetting which becomes significant at about 2~arcmin. Our simplified mirror model also leads to a flat focal plane, while a Wolter-Schwarzschild optic typically has a slightly curved focal surface. Detailed ray-trace models specific to the Athena SPOs are available\cite{10.1117/12.2232230,10.1117/12.2594461,10.1117/12.2628133} and can be coupled with our code for CAT gratings and detectors.

@@

\subsection* {Code, Data, and Materials Availability}
We use the ray-tracing code MARXS\cite{2017AJ....154..243G,marxs2.0}, a Python code developed by us under an open source license. MARXS is available at \url{https://github.com/chandra-marx/marxs}; simulations in this article are done with version 2.0. The database of efficiencies for gratings is considered a trade secret by the Space Nanotechnology Laboratory and cannot be shared in open repositories - contact co-author Ralf Heilmann for access.

@@

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Peter Cheimets}
Peter Cheimets received a BS in Mechanical Engineering in 1978 from MIT, and a MS in Mechanical Engineering with concentrations in dynamics and control systems from Stanford in 1981. He received a management of engineering degree from Tuft in 2007. He has developed astrophysical instrumentation for the Smithsonian Astrophysical Observatory since 1983, placing instruments on TRACE, Hinode, SDO, and IRIS, and is leading the effort to develop an instrument for MUSE.

\vspace{1ex}
\vspace{2ex}\noindent\textbf{Casey DeRoo}
Casey DeRoo received his BA degree in physics and classical studies in 2011 and his PhD in physics from University of Iowa in 2016. He served as an Astrophysicist at the Smithsonian Astrophysical Observatory. He joined the faculty of the University of Iowa as an assistant professor in 2018, and was named a NASA Astrophysics Nancy Grace Roman Technology Fellow in 2020. His research interests include X-ray optics technologies, nanofabrication, and metrology.

@@

%\vspace{1ex}
%\vspace{2ex}\noindent\textbf{Eric D. %Miller}
%Please send me your biography (approx 75 words).

%\vspace{1ex}
%\vspace{2ex}\noindent\textbf{Andrew Ptak}
%Please send me your biography (approx 75 words).

%\vspace{1ex}
%\vspace{2ex}\noindent\textbf{Randall K. Smith}
%Please send me your biography (approx 75 words).
